---
title: "Using Google Earth as an Area of interest picker"
author: "Siddarta Jairam"
date: "`r strftime(Sys.time(),'%A, %B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.width=7.5, fig.height=4,fig.pos='center',echo=TRUE, comment='>',dpi=800)

figList <- list()
tableList <- list()
updateCnt <- function(cnt, nam){
  cnt[[nam]] <- length(cnt) + 1
}
```

Using the raster package to query geospatial data sets is easy. You can use coordinates, a station list, an extent object, the grid row or col and logical subseting.

```{r}
#Various subseting
```

However, what if you want a more complex object to query the data. What if you wanted a non-standard state region, the area between two lakes and a mountain, the area between two amusement parks, or the region in a flood plain between two evacuation centers? Initially, you would think the answer is external data. This could involve changing the projection, changing the grid size, masking some areas or other data manipulations. At the very least you are tying your analysis to another outside data source that could change over time and has a reputation on its own. Let's face it, during this evaluation and manipulation, you are going to have one eye on Google maps anyways. Many times this level of analysis is not even needed with the grid size that you want or the desired accuracy. What you really want is a back of the envelope area of interest that can be plugged in to find initial results and iterated on if the problem requires it. If you only need to highlight a certain area on a map or have a quick hypothesis about an area, it shouldn't require a whole new data set.

So how do you do this back of the envelope picker? First, you would use Google earth to identify the area you want to study. It has lots of guides using various borders, landmarks, natural structures, elevation etc. Then you would select the area using a new polygon. The polygon is defined by its vertices which you can add and move around to fit the area. When you are finished, name it as something memorable, right click its name in the left window and save with "save place as". Here save it as a KML document with the same name as the polygon. This is just an XML document with a useful header and a listing of all the vertices that you created.

Loading this file into R is kind of convoluted. I am sure there is a dedicated package to do this but I am going to use a procedure that uses things that are probably already loaded in this analysis, rgdal and raster.

```{r}
#KML file to raster object.
```

Now that the area is a Raster object with a simple lat/lon projection, subseting and manipulation is easy to do. Now this procedure can easily be automated with a interactive pop up but, if you are as bad with basic geography as I am (or the average american) you have either Google maps, Google Earth or some other maps in front of you anyways. In this workflow, you can also go back and edit the polygons into something more precise if need be.

This procedure is in the middle of analysis techniques in terms of time invested and accuracy as shown below.

1. Extent - When a bounding box is sufficient
2. Google Earth polygon - When it needs to conform to boundaries or landmarks
3. External data - When you need to query the data to get an area

It may seem that out of these three, this procedure needs the most justification. However, the other two can have plenty of hand-waving under the surface. Grid sizes and the complexity of the AOI has an effect on what you pick as well. Documentation goes up dramatically as you go down the list but you will have to do that with whatever you use.

*rmarkdown files for this post can be found on https://github.com/sidjai/blogs*
