---
title: "Vapply and Assertions"
author: "Siddarta Jairam"
date: "`r strftime(Sys.time(),'%A, %B %d, %Y')`"
output: md_document
---

```{r setup, include=FALSE}
source("utils.R")
knitr::opts_chunk$set(fig.width=7.5, fig.height=4,fig.pos='center',echo=TRUE, comment='>',dpi=800)
figList <- list()
tableList <- list()
updateCnt <- function(cnt, nam){
  cnt[[nam]] <- length(cnt) + 1
}
```

Vapply is often the overlooked data analysis tool in the apply family. lapply gives a developer the advantage of a list output while sapply seemingly handles much of the other cases. The others are shown as having very specific use cases that one should only learn if you can't find another tool to do the job. vapply is often introduced after these other functions and is shown to require a much more rigorous implementation. Many don't want the call to have a uniform ouput with a definite size. For those interested in the speed enhancements, many overlook this step and go straight for vectorizing the process.

What is often missed is that vapply exists in the middle of another continuum, interactive debugging and unit tests. Unlike a lot of functions in R, vapply actually returns errors. Additionally, these errors are actually quite informative. This comes from the function's need for an exact class and number of outputs in each iteration. Look at this output:

```{r, results='asis', error=TRUE}

dat <- list(runDoThings = 2,
						runOtherThings = 3,
						runThirdThing = 54,
						runLastThings = 90)
						

tryCatch({
	des <- vapply(names(dat), function(x){
		paste(x, dat[[x]]/2, sep = ': ')
		5
		},'e')
	},
	error= function(cond){message(cond)})


cat(sapply(names(dat), function(x){
	paste(x, dat[[x]]/2, sep = ': ')
	5
}))
```

Moreover the implementation is self-documenting. The desired result is required in the call itself. With a little standardization using already evaluated example objects, or a call to "rep", one can see what you had in mind for that call. Take this for example:

```{r}
des <- vapply(names(dat), function(x){
	paste(x, rnorm(2, dat[[x]]), sep = ': ')
},rep('e', 2))
```

After 6 months, you don't have to remember how "rnorm" works or what exactly was in "dat", the ending tells you the class and number of "des[, "runDoThings"]". If the process changes to make it mistakenly output a larger vector, this will result in an error. Take the sapply equivalent:

```{r}
des <- sapply(names(dat), function(x){
	res <- paste(x, rnorm(3, dat[[x]]), sep = ': ')
	if(class(res) != class('e')){
		stop(sprintf("Error values must be of type %s, but FUN(X[[%s]]) result is type %s", class('e'), x, class('res')))
	}
	if(length(res) != 3){
		stop(sprintf("Error values must be length %f, but FUN(X[[%s]]) result is length %d", 2, x, length(res)))
	}
	return(res)
})
```

Obviously you can use some other exception handling packages to decrease the verbage, but vapply is still doing a lot under the hood. Additionally, vapply makes it easier to change the specification with an expanded function or scope.

The real advantage of this function comes with combining these two types of assertions and being able to give a more complicated call without losing the clean look of vapply:

```{r}
des <- list()
vapply(dat, function(x){
	out <- paste(x, rnorm(3, x), sep = ': ')
	des <- cbind(des, out)
	return(all(nzchar(out)))


},TRUE)
```

Instead of relying on the output, the processing is done inline which looks more natural coming from any programming language that is dominated by loops. The output is than cleared up for a test. The test then checks 3 different things: the character length, the output's class and number. This assertion only takes up one line and looks quite stylish. The result has all the tests, the col and row names if done on two dimensions. Now this has the disadvantage of not throwing an error immedieatly when found, but with this comes the ability to check if the error can occur in other situations.

```{r}
des <- list()
vapply(dat, function(x){
	out <- paste(x, rnorm(3, x), sep = ': ')
	des <- cbind(des, out)
	return(x > 10)


},TRUE)
```

Its easy now to see that the first two runs have a problem when the rest do not. What was different about these runs that raised an error? Trying the pin point the error can lead to expanding the processing to see when it raises and when it does not. In this way, the assertion can grow and accommodate a growing procedure.

This workflow keeps things documented and makes debugging easier. This comes with a drawback of making the call stack a bit confusing. However, the ability to redo the process with other constraints more than makes up for it. To be useful, assertions have to be used often so that the code is separated by gates of things you know to be true. This code makes it easy and fun to add assertions to regular function calls without a bulky if statement with a "stop" call or another package dependency.
